On Optimality of Finite Element Integration

### 2015-12-16 TOMS_Paper/assembly.tex
minor fixes

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item No transformation takes place. In general, one should consider applying generalized code motion or creating a temporary for CCSE -- but it is not the case of this simple example.
\item Spatial sharing is found, so the expression is recast as $f_i (a + c) + g_i (b + d)$. This promotes temporal sharing and, therefore, code motion. 
\end{enumerate}
The reader can verify that option 2) improves the operation count over 1) if and only if the size of $L_i$ is strictly greater than $c=1$. One can appreciate the dichotomy between 1) and 2) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}, obviously with different values of $c$.

We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives. Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations. Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^0, P^1, ..., P^{p-1}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item Calculate the cost of each $P \in \mathbb{P}$, as $c_P = {min(}|P|,\ |S_P|{)}$. Intuitively, $c_P$ indicates the number of operations induced by $P$ at every innermost loop iteration once sharing elimination will be applied. Take $P$ such that $c_P$ is minimum.
%; if multiple partitions have same cost, the preference is for the one with the largest occurrences in the input expression.
\item Choose the optimal strategy between points (i) and (ii), described in Section~\ref{sec:se-rln}. If $|P| < |S_P|$, (i) is applied. If $|P| > |S_P|$, (ii) is applied. Otherwise, $|P| = |S_P|$ means that (i) and (ii) have the same impact on the innermost loop. However, the increase in operation count in the outer loops, $i_l$, may still be different. In particular, if (i) were applied, $i_l = \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$, while if (ii) were applied, $i_l = \sum_{s \in S_P} (n_{s} - 1)$. Note that $n_o$ and $n_s$ capture the number of additions required after factorization of $o$ and $s$. We then select (i) or (ii) such that $i_l$ is minimum. 
\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.
\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above). The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
%TODO : this needs to be applied to all lifted sub-expressions as well
\end{enumerate}
\end{Algo}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$. What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$. In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved. We distinguish two cases.
\begin{itemize}
\item $|P| < |S_P|$. This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place. We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
\item $|P| \geq |S_P|$. The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected. Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion. It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$. Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions). Conversely, if we scheduled $P_{min}$ first, the increase would be -- assuming, without loss of generality, that $|S_{P_{min}}| = n - 1$ -- equal to $2(n - 1) - 1$. 
\end{itemize}

>>>>

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, we take a matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. For simplicity, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions, and consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$. Thus an entry in $y_{L_i}$ is of the form $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression, we basically have two options:
\begin{enumerate}[label=(\roman*)]
\item Applying generalized code motion (or simply creating a temporary for CCSE) -- not necessary in this example since there is just one loop.
\item Searching for spatial sharing, expanding and then factorizing the expression to promote temporal sharing, applying generalized code motion -- the expression in the example would be recast as $f_i (a + c) + g_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) for any size of $L_i$. In general, however, the optimal option depends on both the expansion cost in (ii) and the loop size. One can appreciate the dichotomy between (i) and (ii) in several expressions originating from a wide range of variational forms\footnote{In fact, it systematically appears in all of the forms used for experimentation (Section~\ref{sec:perf-results}), and many more.}. We then regard as \textit{structured tensor operation} any operation between two tensors that result in sub-expressions for which there is no ambiguity in spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of two alternatives.

Of notable importance, in the perspective of achieving quasi-optimality, is the fact that sub-expressions depending on the multilinear loop nest always result from structured tensor operations. Often, structured tensor operations characterize the $\sigma_{ei}$ sub-expressions too. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results} (e.g., the Cauchy-Green tensor).

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into $p$ disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that multilinear operands in $P$ share a set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.
\item Calculate the operation count induced by applying strategy (i) or (ii) -- as described in Section~\ref{sec:se-rln} -- to $P \in \mathbb{P}$. In particular: $c_P^i = |P||\Lambda| + \sum_{o \in P} (c_{o}^{l} + n_{o} - 1)$ and $c_P^{ii} = |S_P||\Lambda| + \sum_{s \in S_P} (n_{s} - 1)$. Note that:
\begin{itemize}
\item $|P||\Lambda|$ ($|S_P||\Lambda|$) represents the operation count in the innermost loop
\item $n_o$ ($n_s$) captures the number of additions once $o$ ($s$) is factorized
\end{itemize}
We define the cost of $P$ as $c_P = {min(}c_P^{i},\ c_P^{ii}{)}$.
\item For all $P \in \mathbb{P}$ such that $c_P = c_P^i$, for all multilinear operands $o \in P$, apply generalized code motion. Update $\mathbb{P}$ by replacing any hoisted multilinear operand with the corresponding temporary symbol.
\item For all $P \in \mathbb{P}$ such that $c_P^{ii} - |S_P||\Lambda| > 0$ -- that is, 
\item Perform a depth-first visit of the expression tree and, for each $L_i$-dependent expression, apply sharing elimination assuming a structured tensor operation is in place (i.e., as in point 4 above). The result may be sub-optimal if this speculative assumption does not hold (i.e., only a subset of code motion opportunities are exposed).
%TODO : this needs to be applied to all lifted sub-expressions as well
\end{enumerate}
\end{Algo}

%If there exist $P^i$, $P^j$, and $P^k$
%\item If $\mathbb{P} \setminus \lbrace P\rbrace$ is not empty, go back to point 2. Otherwise, step to point 6.

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $m_T$ represent the set of tables $T$ resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $m_{T_{1}}$ and $m_{T_{2}}$. If $m_{T_{1}} \cap m_{T_{2}} \neq \emptyset$, at least one table is assigned to the same temporary. $\Gamma$ would then skip a potential global optimum, since $\theta$ only reasons ``locally'' (i.e., on a monomial basis). By assumption, this case is neglected. The assumption is reasonable because such a scenario is of pathological nature (we have never experienced it) and, with high probability, the retrieved solution would be extremely close to the ideal one.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ will be pre-evaluated (as by assumption), there may be a potential gain in sharing elimination that is lost, which potentially leads to a locally optimal solution -- instead of a global one. We prevent this situation by construction. $\Gamma$ models the problem as a graph, analyzing all possible bipartitions that may break the quasi-optimality claim (lines 7-13 in Figure~\ref{algo:gamma}). Since the graph is in practice very small (the number of monomials in a form rarely exceeds a few units), this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of linearity, the operation count induced by a partition $P \in \mathbb{P}$ in the innermost loop is given by $c_P = {min(}|P|,\ |S_P|{)}$. What we have to prove is the fact that by taking each time $P$ such that $c_P$ is minimum -- henceforth, $P_{min}$ -- leads to quasi-optimality. We do it by contradiction. Let us assume that there exists another ordering for $\mathbb{P}$ leading to $c_{\Lambda''_I} < c_{\Lambda'_I}$. In particular, this implies the existence of $P$, with $c_{P} > c_{P_{min}}$, such that if $P$ is scheduled prior to $P_{min}$, a lower operation count is achieved. We distinguish two cases.
\begin{itemize}
\item $|P| < |S_P|$. This is the simplest scenario, since generalized code motion can be applied to the multilinear operands in $P$ and no disruptive re-factorization takes place. We would then have $c_{\Lambda''_I} = c_{\Lambda'_I}$, which clearly contradicts the hypothesis.
\item $|P| \geq |S_P|$. The expression is re-factorized such that the multilinear symbols in $S_P$ can be collected. Assuming $|S_P| = n$, this process exposes $n$ sub-expressions amenable to code motion. It is possible that these $n$ sub-expressions include all of the multilinear operands (symbols) in $P_{min}$. Due to linearity, the innermost loop will suffer from an increase in operation count of $2n - 1$ ($n$ products, $n-1$ additions). Conversely, assuming $|S_{P_{min}}| = n - 1$, if we scheduled $P_{min}$ first, . 
\end{itemize}
